Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2008-12-02 docs/pubs/ISEI6/main.tex
upto attribute blocks

An \owlifier\ table defines an OWL-DL
\cite{smith04:_owl_web_ontol_languag_guide} ontology through a set of
\emph{blocks} representing one or more ontology definitions.  Each
non-empty row in an \owlifier\ table corresponds to a block. The type
of the block is given in the first column of the row. We describe each
type of block supported by \owlifier\ below. Here we assume that if
any properties or classes used in a block are undefined, i.e., are not
imported from another ontology, then they are added to the ontology
being specified via the \owlifier\ table. In general, we name blocks
according to the more generic terms used in
\cite{bowers08:_concep_model_framew_for_expres,madina07:_ontol_for_descr_and_synth}
instead of those of OWL-DL. This allows \owlifier\ to generate
ontologies that extend the observational model of
\cite{madina07:_ontol_for_descr_and_synth} for data annotation, and in
certain cases avoids confusion with established terms used within
biology.

\myblock{Import Blocks.} Import blocks assign namespace labels to
external ontologies. Each external ontology is imported into the
current ontology. We refer to the ontologies of import blocks as
\emph{imported ontologies}.  Using import blocks, classes and
properties of imported ontologies can be used within other blocks of
an \owlifier\ table.  Rows containing import blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline
    \texttt{import} & $n$ & $u$ \\ \hline 
  \end{tabular} 
\end{itemize}
where $n$ is a namespace label and $u$ is an OWL ontology URI. Classes
and properties from imported ontologies are referenced by prefixing
the namespace label $n$ to the corresponding class or property name in
the normal way. The following example block imports the ``earth
realm'' ontology from the set of SWEET ontologies
\cite{raskin:_seman_web_for_earth_and}
\begin{tabbing}
  ~~\texttt{import sweet
    http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
\end{tabbing}
With this import block the class denoting marine ecosystems can be
referred to from within an \owlifier\ table using the expression
\texttt{sweet:MarineEcosystem}. Because this class refers to a class
in another ontology, we refer to it as an \emph{imported class}.

\myblock{Entity Blocks.} Entity blocks introduce new OWL classes and
specify subclass relationships. Imported classes may also be used
within entity blocks by prefixing class names with namespace labels
(see above).  We use the term `entity' to conform to the observational
model of \cite{bowers08:_concep_model_framew_for_expres} and to avoid
confusion with the term 'class' used within biological
taxonomies. Rows containing entity blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{entity} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where each $c$ is a class, and each $c_i$ is asserted in the current
ontology to subsume $c_{i+1}$, for $1 \le i < n$. That is, each $c_i$
in a concept block induces the description-logic axiom $c_{i+1}
\sqsubseteq c_i$.  If both $c_i$ and $c_{i+1}$ are imported classes,
we say that the block defines an ``articulation'' (i.e., mapping)
between the two classes. The following concept block defines a simple
subclass hierarchy
\begin{tabbing}
  ~~\texttt{entity PhysicalFeature AquaticPhysicalFeature River}
\end{tabbing}
stating that Physical Feature, Aquatic Physical Feature, and River are
classes, and that River is a subclass of Aquatic Physical Feature, and
Aquatic Physical Feature is a subclass of Physical Feature.

\myblock{Synonym Blocks.} Synonym blocks define equivalence
relationships between ontology classes.  Rows containing synonym
blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{synonym} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
stating that each class $c_i$ is equivalent to class $c_{i+1}$ in the
current ontology, for $1 \le i < n$. Each $c_i$ in a synonym blocks
induces a description logic axiom of the form $c_i \equiv c_{i+1}$. As
an example, the block
\begin{tabbing}
~~\texttt{synonym Maize Corn}
\end{tabbing}
defines Maize and Corn as synonym (i.e., equivalent) classes.

\myblock{Overlap Blocks.} Except in certain situations (see
\secref{sec:properties}), classes are assumed to be disjoint.  Overlap
blocks explicitly relax this assumption by stating that a set of
classes may have overlapping instances. Rows containing overlap blocks
take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
stating that each class $c_i$ is allowed to share instances with each
class $c_j$, for $1 \le i,j \le n$. In particular, a given $c_i$ and
$c_j$ in an overlap block are not defined to be disjoint classes in
the current ontology. As an example, consider the following entity
blocks that define the classes Estuary, Lagoon, and Marsh as
subclasses of Ecological Habitats.
\begin{tabbing}
\begin{tabular}{lll}
  \texttt{entity} & \texttt{EcologicalHabitat} & \texttt{Estuary} \\ 
  \texttt{entity} & \texttt{EcologicalHabitat} & \texttt{Lagoon} \\ 
  \texttt{entity} & \texttt{EcologicalHabitat} & \texttt{Marsh} \\ 
\end{tabular}
\end{tabbing}
Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
as disjoint classes. To relax this assumption and allow Lagoons to
also be kinds of Estuaries, we explicitly define the overlap block
\begin{tabbing}
\begin{tabular}{lll}
  \texttt{overlap} & \texttt{Estuary} & \texttt{Lagoon}
\end{tabular}
\end{tabbing}
In general, overlap blocks are rarely used but provide a mechanism to
override the default behavior of \owlifier\ in asserting disjoint
classes.

\myblock{Relationship Blocks.} Relationship blocks define required
\emph{object} properties of classes. A property block has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{relationship} & $R$ & $C_1$
    & $C_2$ & \dots & $C_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $R$ is an object property and each $C_i$ is a class, for $1 \le
i \le n$.  For every class $C_i$, the property block induces the DL
axiom \[C_i \sqsubseteq \exists R . C_{i+1} \] stating that each
instance of $C_i$ has, amongst possibly other things, a relationship
through $R$ to some instance of $C_{i+1}$.  The following example
\begin{tabbing}
   ~~need example here
\end{tabbing}

\myblock{Transitive Blocks.} Transitive blocks state that a property
is transitive. That is, if $P$ is transitive and a concept instance
$O_1$ is related to an instance $O_2$ by $P$, and $O_2$ is related to
an instance $O_3$ by $P$, then $O_1$ is also by definition related to
$O_3$ by $P$. A transitive block has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{transitive} & $R$ & $C_1$
    & $C_2$ & \dots & $C_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $P$ is an object property. For example, the block
\begin{quote}\texttt{transitive} \texttt{hasPart} \texttt{Body}
  \texttt{Head} \texttt{Eye} \texttt{Retina}
\end{quote} states that a body has at least one head, a head has at
least one eye, and an eye has at least one retina.

\myblock{Attribute Blocks.} Attribute blocks are used to define the
required \emph{datatype} properties of concepts. An attribute block
has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{attribute} & $P$ & $D$ & $C_1$
    & $C_2$ & \dots & $C_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where $P$ is a datatype property, each $C_i$ is a concept for $1 \le i
\le n$, and $D$ is a datatype (\texttt{anyValueType}, \texttt{string},
\texttt{int}, etc.). For every concept $C_i$, the property block
induces the DL axiom \[C_i \sqsubseteq (\exists P . D)\] stating that
each instance of $C_i$ has, amongst possibly other things, a
relationship through $P$ to a data value of type $D$.

>>>>

An \owlifier\ table defines an OWL-DL
\cite{smith04:_owl_web_ontol_languag_guide} ontology through a set of
\emph{blocks} representing one or more ontology definitions.  Each
non-empty row in an \owlifier\ table corresponds to a block. The type
of the block is given in the first column of the row. We describe each
type of block supported by \owlifier\ below. Here we assume that if
any properties or classes used in a block are undefined, i.e., are not
imported from another ontology, then they are added to the ontology
being specified by the \owlifier\ table. In general, we name blocks
according to the more generic terms used in
\cite{bowers08:_concep_model_framew_for_expres,madina07:_ontol_for_descr_and_synth}
instead of those of OWL-DL. This allows \owlifier\ to generate
ontologies that extend the observational model of
\cite{bowers08:_concep_model_framew_for_expres,madina07:_ontol_for_descr_and_synth}
for data annotation, and in certain cases avoids confusion with
established terms used within biology.

\myblock{Import Blocks.} Import blocks assign namespace labels to
external ontologies. Each external ontology is imported into the
current ontology. We refer to the ontologies of import blocks as
\emph{imported ontologies}.  Using import blocks, classes and
properties of imported ontologies can be used within other blocks of
an \owlifier\ table.  Rows containing import blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline
    \texttt{import} & $n$ & $u$ \\ \hline 
  \end{tabular} 
\end{itemize}
where $n$ is a namespace label and $u$ is an OWL ontology URI. Classes
and properties from imported ontologies are referenced by prefixing
the namespace label $n$ to the corresponding class or property name in
the normal way. As an example, the following block imports the ``earth
realm'' ontology from the set of SWEET ontologies
\cite{raskin:_seman_web_for_earth_and}
\begin{tabbing}
  ~~\texttt{import}  \texttt{sweet} 
  \texttt{http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
\end{tabbing}
With this import block the class denoting Marine Ecosystems (a class
defined in the above ontology) can be referred to from within an
\owlifier\ table using the expression
\texttt{sweet:MarineEcosystem}. Because this class refers to a class
in another ontology, we refer to it as an \emph{imported class}.

\myblock{Entity Blocks.} Entity blocks are the primary blocks used to
define ontologies. An entity block introduces new OWL classes and
specifies subclass relationships. Imported classes may also be used
within entity blocks by prefixing class names with namespace labels
(as described above).  We use the term `entity' to conform to the
observational model of \cite{bowers08:_concep_model_framew_for_expres}
and to avoid confusion with the term `class' used within biological
taxonomies. Rows containing entity blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{entity} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where each $c$ is a class, and each $c_i$ is asserted in the current
ontology to subsume $c_{i+1}$, for $1 \le i < n$. That is, each $c_i$
in a concept block induces the description-logic axiom $c_{i+1}
\sqsubseteq c_i$.  If both $c_i$ and $c_{i+1}$ are imported classes,
we say that the block defines an ``articulation'' (i.e., mapping)
between the two classes. The following concept block defines a simple
subclass hierarchy
\begin{tabbing}
  ~~\texttt{entity} \texttt{PhysicalFeature} 
    \texttt{AquaticPhysicalFeature} \texttt{River}
\end{tabbing}
stating that Physical Feature, Aquatic Physical Feature, and River are
classes, and that River is a subclass of Aquatic Physical Feature, and
Aquatic Physical Feature is a subclass of Physical Feature.

\myblock{Synonym Blocks.} Synonym blocks define equivalence
relationships between ontology classes.  Rows containing synonym
blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{synonym} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
stating that each class $c_i$ is equivalent to class $c_{i+1}$ in the
current ontology, for $1 \le i < n$. Each $c_i$ in a synonym blocks
induces a description logic axiom of the form $c_i \equiv c_{i+1}$. As
an example, the block
\begin{tabbing}
  ~~\texttt{synonym} \texttt{Maize} \texttt{Corn}
\end{tabbing}
defines Maize and Corn as synonyms, i.e., as equivalent classes.

\myblock{Overlap Blocks.} Except in certain situations (see
\secref{sec:properties}), classes are assumed to be disjoint in
\owlifier.  Overlap blocks explicitly relax this assumption by stating
that a set of classes may have overlapping instances. Rows containing
overlap blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
stating that each class $c_i$ is allowed to share instances with each
class $c_j$, for $1 \le i,j \le n$. In particular, a given $c_i$ and
$c_j$ in an overlap block are not defined to be disjoint classes in
the current ontology. As an example, consider the following entity
blocks that define the classes Estuary, Lagoon, and Marsh as
subclasses of Ecological Habitats.
begin
\begin{tabbing}
  ~~\texttt{entity} \texttt{EcologicalHabitat} \texttt{Estuary} \\ 
  ~~\texttt{entity} \texttt{EcologicalHabitat} \texttt{Lagoon} \\ 
  ~~\texttt{entity} \texttt{EcologicalHabitat} \texttt{Marsh} \\ 
\end{tabbing}
Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
as disjoint classes. To relax this assumption and allow, e.g., types
of Lagoons to also be types of Estuaries, we explicitly add the
following overlap block
\begin{tabbing}
  ~~\texttt{overlap} \texttt{Estuary} \texttt{Lagoon}
\end{tabbing}
In general, overlap blocks are rarely used but provide a mechanism to
override the default behavior of \owlifier\ in asserting disjoint
classes.

\myblock{Relationship Blocks.} Relationship blocks define
\emph{required} object properties of classes. An object property
within OWL is a property defined between two class instances. Rows
containing relationship blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{relationship} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c_i$ is a class, for $1 \le
i \le n$.  For every class $c_i$, the relationship block induces the
DL axiom $c_i \sqsubseteq \exists p . c_{i+1}$ stating that each
instance of $c_i$ has, amongst possibly other things, a relationship
through $p$ to some instance of $c_{i+1}$.  For example, the following
block states that the class California Voles live in Grassy Areas.
\begin{tabbing}
  ~~\texttt{relationship} \texttt{livesIn} \texttt{CaliforniaVole} 
    \texttt{GrassyArea}
\end{tabbing}
In some cases, a particular property can apply to a sequence of
classes, and for convenience, each such class can be specified in a
single block. For example, consider the following block.
\begin{tabbing}
  ~~\texttt{relationship} \texttt{directlyBelow} \texttt{Hypolimnion} 
    \texttt{Thermocline} \texttt{Epilimnion} 
\end{tabbing}
This block states that, e.g., within a thermally stratified lake, the
Hypolimnion layer is directly below the Thermocline layer, and the
Thermocline layer is directly below the Epilimnion layer.

\myblock{Transitive Blocks.} Transitive blocks are special cases of
relationship blocks in which the object property is asserted to be
transitive. That is, if $p$ is a transitive property from individuals
$o_1$ to $o_2$ and from $o_2$ to $o_3$, then $p$ is also assumed to be
from $o_1$ to $o_3$. Rows containing transitive blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{transitive} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c_i$ is a class, for $1 \le
i \le n$. As a simple example, the following transitive block
\begin{tabbing}
  ~~\texttt{transitive} \texttt{hasPart} \texttt{Body}
  \texttt{Head} \texttt{Eye} \texttt{Retina}
\end{tabbing} 
states that every Body has a Head as a part, every Head has an Eye as
a part, and every Eye has a Retina as a part. Moreover, because the
part-of property here is defined to be transitive, it is possible to
infer that every Body also has an Eye and a Retina.

\myblock{Attribute Blocks.} Attribute blocks define required
\emph{datatype} properties of concepts. Unlike object properties, a
datatype property within OWL is a property defined between a class
instance and a literal value (e.g., string, integer, etc.). Rows
containing attribute blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{attribute} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where $p$ is a datatype property and $c_i$ is a class, for $1 \le i
\le n$. For each concept $c_i$, the attribute block induces the
description logic axiom $c_i \sqsubseteq \exists p$ stating that each
instance of $c_i$ has, amongst possibly other things, a property $p$
with a literal value. As an example, the following ...
